\Introduction{Introduction} \label{introduction}

\section*{Complex Continuous Optimization and Multi-Agent Systems}

Continuous optimization is a very large field including various methods tailored for diverse specific requirements. While this approach was successful in providing a toolbox of specialized methods, the evolution of industrial needs draws attention to some of its limitations. Indeed, current optimization methods fail to handle the more complex optimization problems. These problems are characterized by heavy calculus, many interdependencies between their components and the diverse expertise domains they involve. Classical optimization tools struggle with these problems because of these factors, and specific methods have been proposed to handle this complexity, giving birth to the field of Multidisciplinary Design Optimization (MDO). However, MDO methods involve possibly important transformations to the original problem in order to divide the problem into simpler ones, which makes this approach somewhat cumbersome and potentially inefficient for highly connected problems.\\
This issue is especially present in the context of complex, real-world system design (aircrafts, space shuttles \emph{etc}, where the complexity of the problem is usually a reflection of the complexity of the system being built. In this context, the problem is not only very large and interconnected but also often not completely defined and is continuously corrected and modified during the design process. The optimization problem is basically in a feedback loop with the designer solving it, where solution of the problem provides new information to the designer, which in return refines the problem formulation and so on. Existing methods are not adapted to such a dynamic process, as they often need to be started from scratch if the problem formulation is modified.

At the same time, new paradigms are being proposed to handle systemic complexity. One of the most successful is the field of Multi-Agent Systems (MAS). This approach proposes to handle problem complexity using systems of interconnected agents. Instead of reducing the problem in order to solve it using a centralized process, MAS techniques preserve the original problem and use decentralized mechanisms in order to spread the solving effort among the agents. MAS has proved successful in the field of combinatorial optimization, on problems such as graph coloring, sensors network or scheduling. However, very few applications of MAS have been proposed in the context of continuous optimization, and these proposals concern specific application domains or optimization categories.

\pagebreak
During the last ten years, the scientific community has relentlessly pursued the effort to bridge the gap between these two apparently irreconcilable approaches: mathematical optimization and MAS. The goal of this thesis is to contribute to this effort by addressing this mostly unexplored potential application field of MAS: complex continuous optimization.

\section*{Contributions of the Thesis}

The first contributions of this thesis concern the design of a MAS-based algorithm for the solving of complex continuous optimization problems:

We study continuous optimization problem and show how all of them share a common structure. Using this observation, \textbf{we propose a representation of continuous optimization problems entities graphs}, which we call Natural Domain Modeling for Optimization (NDMO).

Based on the NDMO representation \textbf{we identify several agent roles for the graph entities. For each agent role we propose a nominal behavior in order to produce a MAS capable of distributing the optimization process}.

In accordance with the AMAS theory, \textbf{we identify a set of Non-Cooperative Situations (NCSs) susceptible to disturb the normal optimization process, and propose a set of cooperation mechanisms to handles them}. This system is not only able to distribute the continuous optimization process, but is also capable of adapting to changes made by the expert to the problem formulation \emph{during solving}.

We demonstrate the modularity of our system by \textbf{introducing additional concerns with the handling of uncertainties propagations}.

From this work follow two more general contributions concerning the design of MAS.

Using the Make Agent Yourself framework \textbf{we propose a component-based architecture for AMAS adapted to the handling of multiple agent roles and NCS-related mechanisms}. This architecture is based on the idea of stackable skills components following the hierarchy of agent roles, providing the correct methods at the required level.

We also provide a more theoretical contribution \textbf{by abstracting the NCS and solving mechanisms into more general Collective Problem Solving Patterns (CPSP)}. These CPSP are based on a more high-level agent role representation, and are abstracted from any direct application domain. They represent specific agent topologies which can be encountered in agent organizations leading to a disruption of the correct system function, as well as of solving mechanisms proposed to handle such configurations. We propose a schematic \enquote{blueprint} representation which synthesize the content of the different patterns.

At last, \textbf{we integrate our MAS into a working prototype and apply it on real-world problems} provided by industrial partners Airbus and Snecma, in the context of the \emph{Integrated Design for Complex Systems} project, a french National Research Agency funded project.

\pagebreak
\section*{Manuscript Organization}
This thesis is divided into 4 parts:
\begin{enumerate}[P{a}rt I.] %the {a} avoids this letter to be used as the counter (the I is used instead)
\item This part introduces the context of the study by presenting an overview of the continuous optimization field, MAS for optimization and the Adaptive Multi-Agent Systems theory.
\item This part presents a MAS for solving continuous optimization problems. We propose a modeling of a continuous optimization problem as an agent graph, and describe some cooperative behaviors for the different agent roles.
\item We present in this part our work on the integration of our MAS into a working prototype. Based on our experience we propose a component based architecture for AMAS agents and extend previous works on providing AMAS engineering tools by introducing CPSP.
\item In this part we present the experiments we did in order to evaluate and validate our approach.
\end{enumerate}